---
output:
  html_document
bibliography: ref.bib
---

# Advanced trajectory-related DE analyses

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

## Overview

The previous sections have focused on a very simple and efficient - but largely effective - approach to trend fitting.
Alternatively, we can use more complex strategies that involve various generalizations to the concept of linear models.
For example, generalized additive models (GAMs) are quite popular for pseudotime-based DE analyses 
as they are able to handle non-normal noise distributions and a greater diversity of non-linear trends.
We demonstrate the use of the GAM implementation from the `r Biocpkg("tradeSeq")` package on the Nestorowa dataset below.
Specifically, we will take a leap of faith and assume that our pseudotime values are comparable across paths of the MST,
allowing us to use the `patternTest()` function to test for significant differences in expression between paths.

```{r}
# Getting rid of the NA's; using the cell weights
# to indicate which cell belongs on which path.
nonna.pseudo <- pathStat(tscan.pseudo)
not.on.path <- is.na(nonna.pseudo)
nonna.pseudo[not.on.path] <- 0
cell.weights <- !not.on.path
storage.mode(cell.weights) <- "numeric"

# Fitting a GAM on the subset of genes for speed.
library(tradeSeq)
fit <- fitGAM(counts(sce.nest)[1:100,], 
    pseudotime=nonna.pseudo,
    cellWeights=cell.weights)

res <- patternTest(fit)
res$Symbol <- rowData(sce.nest)[1:100,"SYMBOL"]
res <- res[order(res$pvalue),]
head(res, 10)
```

From a statistical perspective, the GAM is superior to linear models as the former uses the raw counts.
This accounts for the idiosyncrasies of the mean-variance relationship for low counts and avoids some problems with spurious trajectories introduced by the log-transformation (`r link("normalization-transformation", "OSCA.basic")`).
However, this sophistication comes at the cost of increased complexity and compute time,
requiring parallelization via `r Biocpkg("BiocParallel")` even for relatively small datasets.

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
