---
output:
  html_document
bibliography: ref.bib
---

# Principal curves

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

## Overview

To identify a trajectory, one might imagine simply "fitting" a one-dimensional curve so that it passes through the cloud of cells in the high-dimensional expression space.
This is the idea behind principal curves [@hastie1989principal], a non-linear generalization of PCA where the axes of most variation are allowed to bend.
The curve itself is defined by iteratively mapping cells onto an initial curve, computing a local average of cells at each position, and using that to define the coordinates of a new curve.
By using information from many cells at once, we obtain a stable fit that is robust to noise in single-cell datasets.
Compared to cluster-based MSTs, principal curves provide a smoother trajectory that contains fewer clumps of cells with identical pseudotimes.
In this chapter, we will use the `r Biocpkg("slingshot")` package [@street2018slingshot] to demonstrate the use of principal curves on the Nestorowa dataset [@nestorowa2016singlecell].

```{r, results='asis', echo=FALSE}
extractFromPackage("nestorowa-hsc.Rmd", package="OSCA.workflows",
    chunk="clustering", objects="sce.nest")
```

```{r}
sce.nest
```

## Using `r Biocpkg("slingshot")`

In the simplest setting where no branches are expected, we use `r Biocpkg("slingshot")` to fit a single principal curve to our dataset.
We do so using the low-dimensional PC coordinates for denoising and speed (`r link("dimensionality-reduction", "OSCA.basic")`).
This function yields a pseudotime ordering of cells based on their relative positions when the cells are mapped onto the curve.

```{r}
library(slingshot)
sce.sling <- slingshot(sce.nest, reducedDim='PCA')
head(sce.sling$slingPseudotime_1)
```

We can then visualize the path taken by the fitted curve in any desired space with `embedCurves()`.
For example, Figure \@ref(fig:traj-princurve-tsne-nest) shows the behavior of the principle curve on the $t$-SNE plot.
Note that this may not always yield aesthetically pleasing plots if the $t$-SNE algorithm decides to arrange clusters so that they no longer match the ordering of the pseudotimes.

```{r traj-princurve-tsne-nest, fig.cap="$t$-SNE plot of the Nestorowa HSC dataset where each point is a cell and is colored by the _slingshot_ pseudotime ordering. The fitted principal curve is shown in black."}
embedded <- embedCurves(sce.sling, "TSNE")
embedded <- slingCurves(embedded)[[1]] # only 1 path.
embedded <- data.frame(embedded$s[embedded$ord,])

library(scater)
plotTSNE(sce.sling, colour_by="slingPseudotime_1") +
    geom_path(data=embedded, aes(x=Dim.1, y=Dim.2), size=1.2)
```

## Handling branched trajectories

The previous call to `slingshot()` assumed that all cells in the dataset were part of a single curve.
To accommodate more complex events like bifurcations, we use our previously computed cluster assignments to build a rough sketch for the global structure in the form of a MST across the cluster centroids.
Each path through the MST from a designated root node is treated as a lineage that contains cells from the associated clusters.
Principal curves are then simultaneously fitted to all lineages with some iterative averaging across curves to encourage consistency in shared clusters across lineages.
This process yields a matrix of pseudotimes where each column corresponds to a lineage and contains the pseudotimes of all cells assigned to that lineage.

```{r}
sce.sling2 <- slingshot(sce.nest, cluster=colLabels(sce.nest), reducedDim='PCA')
pseudo.paths <- slingPseudotime(sce.sling2)
head(pseudo.paths)
```

By using the MST as a scaffold for the global structure, `slingshot()` can accommodate branching events based on divergence in the principal curves (Figure \@ref(fig:traj-princurve-clustered-nest)).
However, unlike `r Biocpkg("TSCAN")`, the MST here is only used as a rough guide and does not define the final pseudotime.

```{r traj-princurve-clustered-nest, fig.cap="UMAP plot of the Nestorowa HSC dataset where each point is a cell and is colored by the average _slingshot_ pseudotime across paths. The principal curves fitted to each lineage are shown in black."}
sce.nest <- runUMAP(sce.nest, dimred="PCA")
reducedDim(sce.sling2, "UMAP") <- reducedDim(sce.nest, "UMAP")

# Taking the rowMeans just gives us a single pseudo-time for all cells. Cells
# in segments that are shared across paths have similar pseudo-time values in 
# all paths anyway, so taking the rowMeans is not particularly controversial.
shared.pseudo <- rowMeans(pseudo.paths, na.rm=TRUE)

# Need to loop over the paths and add each one separately.
gg <- plotUMAP(sce.sling2, colour_by=I(shared.pseudo))
embedded <- embedCurves(sce.sling2, "UMAP")
embedded <- slingCurves(embedded)
for (path in embedded) {
    embedded <- data.frame(path$s[path$ord,])
    gg <- gg + geom_path(data=embedded, aes(x=Dim.1, y=Dim.2), size=1.2)
}

gg
```

We can use `slingshotBranchID()` to determine whether a particular cell is shared across multiple curves or is unique to a subset of curves (i.e., is located "after" branching).
In this case, we can see that most cells jump directly from a global common segment (`1,2,3`) to one of the curves (`1`, `2`, `3`) without any further hierarchy, i.e., no noticeable internal branch points.

```{r}
curve.assignments <- slingBranchID(sce.sling2)
table(curve.assignments)
```

```{r, echo=FALSE}
# Sanity checks.
tab <- table(curve.assignments)
stopifnot(max(tab)==tab["1,2,3"])
stopifnot(sum(tab[c("1,2,3", "1", "2","3")]) > 0.95)
```

## Dealing with large datasets

For larger datasets, we can speed up the algorithm by approximating each principal curve with a fixed number of points.
By default, `slingshot()` uses one point per cell to define the curve, which is unnecessarily precise when the number of cells is large.
Applying an approximation with `approx_points=` reduces computational work without any major loss of precision in the pseudotime estimates.

```{r}
sce.sling3 <- slingshot(sce.nest, cluster=colLabels(sce.nest), 
    reducedDim='PCA', approx_points=100)
pseudo.paths3 <- slingPseudotime(sce.sling3)
head(pseudo.paths3)
```

```{r, echo=FALSE}
stopifnot(median(pseudo.paths3 - pseudo.paths, na.rm=TRUE) < 0.1)
```

## Adding an OMEGA cluster 

The MST can also be constructed with an OMEGA cluster to avoid connecting unrelated trajectories.
This operates in the same manner as (and was the inspiration for) the outgroup for `r Biocpkg("TSCAN")`'s MST.
Principal curves are fitted through each component individually, 
manifesting in the pseudotime matrix as paths that do not share any cells.

```{r traj-princurve-omag-nest, fig.cap="UMAP plot of the Nestorowa HSC dataset where each point is a cell and is colored by the average _slingshot_ pseudotime across paths. The principal curves (black lines) were constructed with an OMEGA cluster."}
sce.sling4 <- slingshot(sce.nest, cluster=colLabels(sce.nest), 
    reducedDim='PCA', approx_points=100, omega=TRUE)
pseudo.paths4 <- slingPseudotime(sce.sling4)
head(pseudo.paths4)

shared.pseudo <- rowMeans(pseudo.paths, na.rm=TRUE)
gg <- plotUMAP(sce.sling4, colour_by=I(shared.pseudo))
embedded <- embedCurves(sce.sling4, "UMAP")
embedded <- slingCurves(embedded)
for (path in embedded) {
    embedded <- data.frame(path$s[path$ord,])
    gg <- gg + geom_path(data=embedded, aes(x=Dim.1, y=Dim.2), size=1.2)
}
gg
```

```{r, echo=FALSE}
# Sanity check.
stopifnot(all(is.na(pseudo.paths4[,1]) | is.na(pseudo.paths4[,3])))
stopifnot(all(is.na(pseudo.paths4[,2]) | is.na(pseudo.paths4[,3])))
```

## Further comments

The use of principal curves adds an extra layer of sophistication that complements the deficiencies of the cluster-based MST.
The principal curve has the opportunity to model variation within clusters that would otherwise be overlooked;
for example, `r Biocpkg("slingshot")` could build a trajectory out of one cluster while `r Biocpkg("TSCAN")` cannot.
Conversely, the principal curves can "smooth out" circuitous paths in the MST for overclustered data,
ignoring small differences between fine clusters that are unlikely to be relevant to the overall trajectory.

That said, the structure of the initial MST is still fundamentally dependent on the resolution of the clusters.
One can arbitrarily change the number of branches from `r Biocpkg("slingshot")` by tuning the cluster granularity,
making it difficult to use the output as evidence for the presence/absence of subtle branch events.
If the variation within clusters is uninteresting, the greater sensitivity of the curve fitting to such variation may yield irrelevant trajectories where the differences between clusters are masked.
Moreover, `r Biocpkg("slingshot")` is no longer obliged to separate clusters in pseudotime,
which may complicate intepretation of the trajectory with respect to existing cluster annotations.

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
