---
output:
  html_document
bibliography: ref.bib
---

# Detecting DE genes along trajectories

```{r setup, echo=FALSE, results="asis"}
library(rebook)
chapterPreamble()
```

## Overview

Once we have constructed a trajectory, the next step is to characterize the underlying biology based on its DE genes.
The aim here is to find the genes that exhibit significant changes in expression across pseudotime,
as these are the most likely to have driven the formation of the trajectory in the first place.
The overall strategy is to fit a model to the per-gene expression with respect to pseudotime,
allowing us to obtain inferences about the significance of any association.
We can then prioritize interesting genes as those with low $p$-values for further investigation.

In this chapter, we will fit linear model to the log-expression values with respect to the pseudotime.
We can then apply standard statistical techniques to test for genes with significant trends.
Despite its simplicity, this approach is quite effective at generating a sensible ranking across genes.
We demonstrate on the @nestorowa2016singlecell dataset with the `r Biocpkg("TSCAN")` pseudotimes from Figure \@ref(fig:tscan-nest-omega).

```{r, results='asis', echo=FALSE}
extractFromPackage("nestorowa-hsc.Rmd", package="OSCA.workflows",
    chunk="clustering", objects="sce.nest")
```

```{r}
sce.nest

library(TSCAN)
quick.pseudo <- quickPseudotime(sce.nest, use.dimred="PCA", outgroup=TRUE)
tscan.pseudo <- pathStat(quick.pseudo$ordering)
```

```{r, echo=FALSE}
stopifnot(igraph::components(quick.pseudo$mst)$no == 2)

# First path is the one that lives by itself.
stopifnot(colnames(tscan.pseudo)[1]=="7")
stopifnot(which.max(-colSums(!is.na(tscan.pseudo)))==1)
```

## Changes along a path

```{r, echo=FALSE}
target <- "6"
```

We use the `testPseudotime()` utility to fit a natural spline to the expression of each gene against one of the pseudotime vectors - in this case, the path ending in cluster `r target`.
A spline is useful as it allows us to model a range of non-linear relationships in the data.
We then perform an analysis of variance (ANOVA) to determine if any of the spline coefficients are significantly non-zero,
i.e., there is some significant trend with respect to pseudotime.

```{r}
chosen <- "6"
pseudo <- testPseudotime(sce.nest, pseudotime=tscan.pseudo[,chosen])
pseudo$SYMBOL <- rowData(sce.nest)$SYMBOL
sorted <- pseudo[order(pseudo$p.value),]
sorted
```

```{r, echo=FALSE}
stopifnot(identical(chosen, target))
```

In practice, it is helpful to pair the spline-based ANOVA results with a fit from a much simpler model
where we assume that there exists a linear relationship between expression and the pseudotime.
This yields an interpretable summary of the overall direction of change in the `logFC` field above,
complementing the more poweful spline-based model used to populate the `p.value` field.
In contrast, the magnitude and sign of the spline coefficients cannot be easily interpreted.

Examination of the top downregulated genes suggests that this pseudotime represents a transition away from myeloid identity,
based on the decrease in expression of genes such as _Mpo_ and _Plac8_ (Figure \@ref(fig:nest-1-simple-down)).

```{r nest-1-simple-down, fig.asp=3, fig.width=6, fig.cap="Expression of the top 10 genes that decrease in expression with increasing pseudotime along the first path in the MST of the Nestorowa dataset. Each point represents a cell that is mapped to this path and is colored by the assigned cluster."}
up.left <- sorted[sorted$logFC < 0,]
head(up.left, 10)

library(scater)
best <- head(up.left$SYMBOL, 10)
sce.nest$TSCAN.path <- tscan.pseudo[,chosen]
plotExpression(sce.nest, features=best, swap_rownames="SYMBOL",
    x="TSCAN.path", colour_by="label")
```

```{r, echo=FALSE, results="hide"}
stopifnot(all(c("Mpo", "Plac8") %in% head(up.left$SYMBOL, 10)))
```

Conversely, the later parts of the pseudotime may correspond to a more stem-like state based on upregulation of genes like _Hlf_.
There is also increased expression of genes associated with the lymphoid lineage (e.g., _Ltb_),
consistent with reduced commitment to the myeloid lineage at earlier pseudotime values. 

```{r nest-1-simple-up, fig.asp=3, fig.width=6, fig.cap="Expression of the top 10 genes that increase in expression with increasing pseudotime along the first path in the MST of the Nestorowa dataset. Each point represents a cell that is mapped to this path and is colored by the assigned cluster."}
up.right <- sorted[sorted$logFC > 0,]
head(up.right, 10)

best <- head(up.right$SYMBOL, 10)
plotExpression(sce.nest, features=best, swap_rownames="SYMBOL",
    x="TSCAN.path", colour_by="label")
```

```{r, echo=FALSE, results="hide"}
stopifnot(all(c("Hlf", "Ltb") %in% head(up.right$SYMBOL, 10)))
```

Alternatively, we can use a heatmap to provide a more compact visualization (Figure \@ref(fig:nest-1-simple-up-heat)).

```{r nest-1-simple-up-heat, fig.width=6, fig.height=7, fig.cap="Heatmap of the expression of the top 50 genes that increase in expression with increasing pseudotime along the first path in the MST of the Nestorowa HSC dataset. Each column represents a cell that is mapped to this path and is ordered by its pseudotime value."}
on.first.path <- !is.na(sce.nest$TSCAN.path)
plotHeatmap(sce.nest[,on.first.path], order_columns_by="TSCAN.path", 
    colour_columns_by="label", features=head(up.right$SYMBOL, 50),
    center=TRUE, swap_rownames="SYMBOL")
```

## Changes between paths

A more advanced analysis involves looking for differences in expression between paths of a branched trajectory.
This is most interesting for cells close to the branch point between two or more paths where the DE analysis may identify genes responsible for the branching event.
The general strategy here is to fit one trend to the unique part of each path immediately following the branch point, followed by a comparison of the fits between paths.

To this end, one particularly tempting approach is to perform another ANOVA with our spline-based model and test for significant differences in the spline parameters between paths.
Unfortunately, a pseudotime value in one path of the MST does not, in general, have any relation to the same value in another path; 
the pseudotime can be arbitrarily "stretched" by factors such as the magnitude of DE or the density of cells, depending on the trajectory reconstruction algorithm.
This compromises any quantitative comparison of trends in expression as we cannot reliably say that they are being fitted to comparable $x$-axes. 

```{r, echo=FALSE, eval=FALSE}
# Here's an example for y'all. We consider a branch from clusters 1 to 2/3.
# Gene A increases in expression at the same rate in both paths; gene B
# increases in only one path, introducing a change in pseudotime magnitude.
clusters <- gl(3, 1000)
mat <- cbind(
    GeneA=c(integer(1000), 1:1000/500, 1:1000/500),
    GeneB=c(integer(1000), 1:1000/100, integer(1000))
)
mat <- mat + rnorm(length(mat), sd=0.1)

library(slingshot)
test <- slingshot(mat, cluster=clusters, approx_points=100, start.clus=1)
test.paths <- slingPseudotime(test)

# Note the differences in the gradients:
(fit1 <- lm(mat[clusters==2,1] ~ 0 + test.paths[clusters==2,1]))
(fit2 <- lm(mat[clusters==3,1] ~ 0 + test.paths[clusters==3,2]))
```

Rather, we employ the much simpler _ad hoc_ approach of fitting a spline to each trajectory and comparing the sets of DE genes.
To demonstrate, we focus on the cluster containing the branch point in the Nestorowa-derived MST (Figure \@ref(fig:tscan-nest-pseudo)).
We recompute the pseudotimes so that the root lies at the cluster center, allowing us to detect genes that are associated with the divergence of the branches.

```{r}
starter <- guessMSTRoots(quick.pseudo$mst, method="minstep")
starter
tscan.pseudo2 <- orderCells(cellData(quick.pseudo$ordering),
    quick.pseudo$mst, start=starter)
```

```{r, echo=FALSE}
stopifnot("3" %in% starter)
stopifnot(igraph::degree(quick.pseudo$mst, "3") == 3)
```

We visualize the reordered pseudotimes using only the cells in our branch point cluster (Figure \@ref(fig:nest-pseudo-reordered)), which allows us to see the correspondence between each pseudotime to the projected edges of the MST.
A more precise determination of the identity of each pseudotime can be achieved by examining the column names of `tscan.pseudo2`, which contains the name of the terminal node for the path of the MST corresponding to each column.

```{r nest-pseudo-reordered, fig.width=10, fig.height=5, fig.cap=sprintf("_TSCAN_-derived pseudotimes around cluster %s in the Nestorowa HSC dataset. Each point is a cell in this cluster and is colored by its pseudotime value along the path to which it was assigned. The overlaid lines represent the relevant edges of the MST.", starter)}
# Making a copy and giving the paths more friendly names.
sub.nest <- sce.nest
sub.nest$TSCAN.first <- pathStat(tscan.pseudo2)[,"4"]
sub.nest$TSCAN.second <- pathStat(tscan.pseudo2)[,"6"]
sub.nest$TSCAN.third <- pathStat(tscan.pseudo2)[,"8"]

# Subsetting to the desired cluster containing the branch point.
keep <- colLabels(sce.nest) == "3"
sub.nest <- sub.nest[,keep]

# Showing only the lines to/from our cluster of interest.
line.data <- quick.pseudo$connected$TSNE
line.data.sub <- line.data[grepl("^3--", line.data$edge) | grepl("--3$", line.data$edge),]
ggline <- geom_line(data=line.data.sub, mapping=aes(x=dim1, y=dim2, group=edge))

gridExtra::grid.arrange(
    plotTSNE(sub.nest, colour_by="TSCAN.first") + ggline,
    plotTSNE(sub.nest, colour_by="TSCAN.second") + ggline,
    plotTSNE(sub.nest, colour_by="TSCAN.third") + ggline,
    ncol=3
)
```

We then apply `testPseudotime()` to each path involving our branch point cluster.
Because we are operating over a relatively short pseudotime interval, we do not expect complex trends and so we set `df=1` (i.e., a linear trend) to avoid problems from overfitting. 

```{r}
pseudo1 <- testPseudotime(sub.nest, df=1, pseudotime=sub.nest$TSCAN.first)
pseudo1$SYMBOL <- rowData(sce.nest)$SYMBOL
pseudo1[order(pseudo1$p.value),]

pseudo2 <- testPseudotime(sub.nest, df=1, pseudotime=sub.nest$TSCAN.second)
pseudo2$SYMBOL <- rowData(sce.nest)$SYMBOL
pseudo2[order(pseudo2$p.value),]

pseudo3 <- testPseudotime(sub.nest, df=1, pseudotime=sub.nest$TSCAN.third)
pseudo3$SYMBOL <- rowData(sce.nest)$SYMBOL
pseudo3[order(pseudo3$p.value),]
```

We want to find genes that are significant in our path of interest (for this demonstration, the third path reported by `r Biocpkg("TSCAN")`) and are not significant and/or changing in the opposite direction in the other paths.
We identify non-significant genes as those with raw $p$-values greater than a threshold of 0.05;
we deliberately omit the multiple testing correction to obtain a more stringent filter for _non_-DE genes.

```{r}
only3 <- pseudo3[which(pseudo3$FDR <= 0.05 & 
    (pseudo2$p.value >= 0.05 | sign(pseudo1$logFC)!=sign(pseudo3$logFC)) &
    (pseudo2$p.value >= 0.05 | sign(pseudo2$logFC)!=sign(pseudo3$logFC))),]
only3[order(only3$p.value),]
```

We observe upregulation of interesting genes such as _Gata2_ and _Apoe_ along with downregulation of _Flt3_ (Figure \@ref(fig:nest-3-versus)).
One might speculate that this path leads to a less differentiated HSC state compared to the other directions.

```{r nest-3-versus, fig.asp=0.5, fig.width=12, fig.cap="$t$-SNE plots of cells in the cluster containing the branch point of the MST in the Nestorowa dataset. Each point is a cell colored by the expression of a gene of interest and the relevant edges of the MST are overlaid on top."}
gridExtra::grid.arrange(
    plotTSNE(sub.nest, colour_by="Flt3", swap_rownames="SYMBOL") + ggline,
    plotTSNE(sub.nest, colour_by="Apoe", swap_rownames="SYMBOL") + ggline,
    plotTSNE(sub.nest, colour_by="Gata2", swap_rownames="SYMBOL") + ggline,
    ncol=3
)
```

```{r, echo=FALSE}
stopifnot(all(c("Flt3", "Gata2", "Apoe") %in% only3$SYMBOL))
```

While simple and practical, this comparison strategy is even less statistically defensible than usual.
The differential testing machinery is not suited to making inferences on the absence of differences,
so technically, we should not have used the non-significant genes to draw any conclusions.
Another limitation is that this approach cannot detect differences in the magnitude of the gradient of the trend between paths; 
a gene that is significantly upregulated in each of two paths but with a sharper gradient in one of the paths will not be DE.
(Of course, this is only a limitation if the pseudotimes were comparable in the first place.)

## Further comments

When a trajectory consists of a series of clusters (as in the Nestorowa dataset),
pseudotime-based DE tests can be considered a continuous generalization of cluster-based marker detection.
One would expect to identify similar genes by performing an ANOVA on the per-cluster expression values,
and indeed, this may be a more interpretable approach as it avoids imposing the assumption that a trajectory exists at all.
The main benefit of pseudotime-based tests is that they encourage expression to be a smooth function of pseudotime,
assuming that the degrees of freedom in the trend fit prevents overfitting.
This smoothness reflects an expectation that changes in expression along a trajectory should be gradual. 

The magnitudes of the $p$-values reported here should be treated with some skepticism.
The same fundamental problems discussed in `r link("p-value-invalidity", "OSCA.basic")` remain;
the $p$-values are computed from the same data used to define the trajectory,
and there is only a sample size of 1 in this analysis regardless of the number of cells.
Nonetheless, the $p$-value is still useful for prioritizing interesting genes 
in the same manner that it is used to identify markers between clusters.

## Session Info {-}

```{r sessionInfo, echo=FALSE, results='asis'}
prettySessionInfo()
```
